A standard for the scholarly citation of archaeological data as an incentive to data sharing

### 2017-11-29 analysis/paper/paper.Rmd
use custom render template & fn to get SAA-style cover page, change spelling to US

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
keywords          : ""
wordcount         : '`r  unname(stringi::stri_stats_latex(readLines("paper.Rmd"))[4])` (includes reference list )'

@@

output:
  papaja::apa6_word
   # :
   # latex_engine: xelatex

abstract: |
  "How do archaeologists share their research data, if at all? We review what data are, according to current influential definitions, and previous work on the benefits, costs and norms of data sharing in the sciences broadly. To understand data sharing in archaeology, we present the results of three pilot studies: requests for data by email; review of data availability in published articles, and analysis of archaeological datasets deposited in repositories. We find that archaeologists are often willing to share, but discipline-wide sharing is patchy and ad hoc. Legislation and mandates are effective at increasing data-sharing, but editorial policies at journals lack adequate enforcement. Although most of data available at repositories are licensed to enable flexible reuse, only a small proportion of the data are stored in structured formats for easy reuse. We present some suggestions for improving the state of date sharing in archaeology, among these is a standard for citing data sets to ensure that researchers making their data publicly available receive appropriate credit."
---

```{r, setup2, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  eval = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

Data are the building blocks of research, and access to data is necessary to fully understand and extend the work of others (Fienberg et al 1985). Yet archaeological data are often not available for a variety of reasons, including a perceived lack of credit for making data available. The goal of this paper is to present a standard for the scholarly citation of archaeological data, akin to the way we cite publications. The need for this has been motivated by Huggert’s (2017) findings of low levels of data citation and reuse in archaeology, and Clarke’s (2015) analysis that digital data preservation and management systems are under-developed in archaeology, despite our obligations as stewards of cultural heritage. A standard for data citation is important because it gives credit to the authors or compilers of the data, and so accords them visibility and recognition for their scholarly contribution in a way that is valued broadly in the discipline. A citation standard is also important for tracking the use of the data, and generating metrics about the impact of the dataset on the discipline. Understanding the use of resources such as datasets is important for allocating resources within archaeology and identifying novel research directions.

Archaeologists work with data in research, commercial, government, and other contexts. A full investigation of data in each of these contexts is beyond the scope of this paper, so we limit this study to what we are most famliar with, the research context. However, many of the practical details we discuss here will also be relevant to archaeologists working in other contexts. We begin by defining what we mean by ‘data’, and reviewing previous work on researchers’ attitudes to sharing data.  We motivate data sharing and citing behaviours by explaining why data sharing is beneficial to archaeology as a discipline, and to individual researchers. Then we survey some of the data sharing and citing norms that have emerged in other research areas. From these norms we derive a standard suitable for citation of archaeological data. Finally, we present some basic instructions on how to share archaeological data.

@@

Defining what data are is an ongoing research problem for information scientists (Leonelli 2016; Borgman 2012). Table \@ref(tab:summary-of-defs-of-data) summarises some of the definitions debated in the philosophy of science literature. A more concrete starting point is this definition from the National Academies of Sciences:

@@

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different challenges and requirements for access, archiving, and so on. The categories shown in Table \@ref(tab:types-of-data) are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: "What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models." These definitions in policy are consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependent on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). How can these definitions and philosophical discussions inform our practice? They tell us that what counts as data is emergent from our professional interactions. Data are the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email. Ther are not exclusively objective archaeometric records, but include interpretive and derivative results. In short, data are anything another person needs to understand our work better.

To translate this into a more applied context, our practice of data sharing it to make available the computer files (e.g. containing text, raw structured data, images, etc.) that we analyse or otherwise use to create the figures, tables and quantitative results that we include in our reports and publications. These include our derived results and secondary data, such as taxonomic and typological classifications. This selection omits materials that could also be considered data according to the definitions noted above (e.g. field notes), but we feel our selection represents the most information-dense and meaningful items that are useful for our community of interest. These are also the items that have been most frequently requested from us by others in our professional interactions. Our citation standard is inclusive and imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion.

@@

While public and community benefits and costs are well-rehearsed and widely agreed on (Tables \@ref(tab:sharing-costs) and \@ref(tab:sharing-benefits)), they offer no, or only weak, tangible incentives to the individual researcher (Rowhani-Farid et al. 2017). Previous work has identified citation advantages as a tangible benefit to individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table \@ref(tab:indivbenefits) for a summary). Increased publication productivity also appeared to be associated with data sharing behaviours. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only five for projects without archived data.

```{r sharing-costs}
tbl <- readr::read_csv("../figures/tbl_reasons_not_to_share_data.csv")
knitr::kable(tbl, caption = "Summary of public and community costs of data sharing, from Stodden (2010) and Borgman (2012)")
```

```{r sharing-benefits}
tbl <- readr::read_csv("../figures/tbl_reasons_to_share_data.csv")
knitr::kable(tbl, caption = "Summary of public and community benefits of data sharing, from Stodden (2010) and Borgman (2012)")
```

@@

Stodden (2010) surveyed registrants at a top machine learning conference (the Neural Information Processing Systems (NIPS) conference) to identify the factors that affect researchers’ decisions to reveal data (Table \@ref(tab:costsbenefits)). Stodden found that communitarian factors influence scientists to share their data, but private incentives determine when scientists choose not to share their code and data. In the top three reasons not to share, we see a lack of incentives to invest time in data sharing (however, there are many other time-consuming tasks in research, and they get done because clear reward structures are in place); a lack of faith in data citation (which we hope this paper will help with); and a lack of an efficient copyright process. Stodden’s (2009a, b) ‘Reproducible Research Standard’, summarised in Table \@ref(tab:stodden-rrs), is a convenient and robust copyright schema that is in wide use for many typical research projects.

@@

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how do these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviours and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). This intensive-restricted spectrum also maps on to Sawyer's (2008) analysis of data sharing in data-rich fields compared to data-poor fields. Data rich fields, such as astronomy, oceanography, and ecology, are have three common characteristics: pooling and sharing data resources is expected; the cumulative nature of work that builds on common data and common methods reduces the likelihood of alternative data sources and other methods being common; and the number of theoretical choices is limited because there is enough data to adequately test competing theories. Data-poor fields have three common characteristics: data are rare, or hard to get, or both; the types of data available often dictate the methods taken; and theories are relatively easy to generate and more difficult to validate. In data-poor fields, researchers often invest substantial time and resources on gathering their own data. The getting of data, and the quality of the data collected, is often a key determinant of the status of a researcher, leading to hoarding of data as an asset to be traded for professional benefit.

Archaeologists can be found all over this spectrum because of the diversity of archaeological research. However, our observations of data sharing in archaeology, and the nature of archaeological research generally, is that it is, for the most part, currently a restricted data-sharing and a data-poor field. There are few standardised and widely used methods for sharing and citing data, a challenge for archaeological data sharing that we hope this paper will help to address.

Even though there is a wide spectrum of data sharing behaviours, overall rates of data sharing by individual researchers are very low in many fields. Wicherts et al. (2006) emailed authors of 141 empirical articles recently published by the American Psychological Association (APA) journals to request a copy of the original data reported in the articles. They received 64 of the 249 datasets they requested, resulting in a data sharing rate of about 26%, despite all the authors of the articles in their study being bound to the APA's Ethical Principles, which state authors must ‘not withhold the data on which their conclusions are based from other competent professionals who seek to verify the substantive claims’ (American Psychological Association, 2001, p. 396). Valpaemel et al. (2015) similarly requested data from 394 papers, published in all issues of four APA journals in 2012 and found that 38% of authors responded with their data. They concluded that although the data sharing rate increased as compared to the study by Wicherts et al., their findings are ‘worrisome’ and indicate a 'poor availability of data'.

@@

The absence of a standardised and sustainable approach to data sharing is starkly demonstrated by Vines et al.’s (2014) study of data availability over time in biology publications. They examined the availability of data from 516 studies of plants or animal morphology between 2 and 22 years old. Vines et al. found that odds of a data set referred to in a publication being available fell by 17% per year. They found that the key factors responsible for the unavailability of data was that the data were lost or on inaccessible storage media.

@@

Here we present the results of three small pilot studies aimed at understanding how archaeologists share their data. First, we emailed a sample of authors of research publications to request access to the data behind their publication. Second, we surveyed a sample of journal articles to see what data was available. Third, we analysed information about archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The small sample sizes of the first two of these pilot studies limit how generalisable the results are, but as the first of their kind for archaeology, they are valuable for proposing hypotheses for larger-scale investigations.

@@

Our sample focused on _Journal of Archaeological Science_ articles on experiments with stone artefacts that were published during `r min(email_survey$'publication year', na.rm = TRUE)` and `r max(email_survey$'publication year', na.rm = TRUE)`. We focused on stone artefact experiments because all research materials in these projects are generated by the researcher, so there are no other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. Because we contacted authors of papers at the intersection of (1) containing eminently sharable data and (2) our _bona fide_ research interests (i.e. stone artefacts for BM), this is a targeted, non-random sample that may not be representative of the entire archaeological community. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymised so the specific authors we contacted are not identifiable.

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and received `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We received `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing.

Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014). Our finding is not unique - Chambers' (2017) examined 50 PLOS One papers on brain images published in November 2014 - May 2015 and found that only 38% of these papers archived their data in an open repository. Similarly, Nuijten et al. (2017) found that of 462 PLOS One articles that promised data availability, 29% of these did not have any available data. Unfulfilled promises to share data are not unique to PLOS One (Kidwell et al. 2016), but these studies highlight the need to resolve ambiguity among authors about what counts as data, and for editors and peer reviwers to more critically assess whether an author has made their data openly available.

We see similar rates of sharing in response to email requests among archaeologists to what have been described in other disciplines (Table \@ref(tab:sharing-rates), Tenopir et al. 2015; Vanpaemel et al. 2015). However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses we received suggests two key reasons why authors are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data (a typical concern of researchers in data-poor fields), and that their data are not in a form suitable for sharing and the task of organising those data is not a high priority. These reasons not to share have also been documented in other research areas, noted above. The absence of patterns in the date of the paper, student status of the author and other variables (Figure \@ref(fig:email-figs)) may relate to more to our small sample size than a real absence of relationships in data sharing behaviour.

@@

```{r student, fig.cap="No relationship between student status and data sharing behaviour"}
sharing_by_student <-
email_survey %>%
  filter(!is.na( `email sent date`)) %>%
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`),
  "yes", "no")) %>%
  group_by(
  `author is graduate student at time of publication?`,
  `content recieved date`) %>%
  tally() %>%
  filter(!is.na(`author is graduate student at time of publication?`)) %>%
  ggplot(aes(`author is graduate student at time of publication?`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3)],
                    guide = FALSE) +
  scale_y_continuous("Number of papers",
                     breaks= pretty_breaks()) +
    xlab("Author was graduate student at time of publication?") +
    theme_bw() +
  coord_fixed(ratio = 0.05)
```

@@

```{r journals, fig.cap="No relationship between sharing behaviour and journal."}

@@

In addition to our email survey, we also conducted a pilot study of a random sample of `r nrow(jas_article_survey)` 48 articles published during February-May 2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviours. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data) (Figure \@ref(fig:data-types-and-files)). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, an open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

The _Journal of Archaeological Science_ has had a 'data disclosure' policy since at least 2013 (Editors, 2013) which states that "All data relating to the article must be made available in Supplementary files or deposited in external repositories and linked to within the article". The data availability rate of `r data_are_avail_perc`% reported in our pilot study reflects either weak enforcement of this policy or an incomplete understanding among editors, reviewers, and authors of how to interpret and implement this policy. The prominence of compositional and dating data among the shared data types suggests a least effort strategy by authors sharing data that do not require extensive cleaning and tidying (because the data are generated by instruments rather than entered by hand, such as artefact attribute data). The size of compositional and dating datasets means that they mostly fit easily into tables in the text or supplementary files, and do not require much effort to organise (e.g. very little metadata is required for others to use these data). The small proportion of papers in our sample using or sharing a scripting language suggests that unlike other fields (Eglen et al. 2017), archaeologists do not appear to recognise code for data analysis as a primary research object to share with the research community. Code for even the simplest operations should be shared along with data because code communicates decisions made about the data analysis that can help others understand, analyse, and reuse the data.

@@

plot_datacentre_rights_table <-
datacentre_rights_table[1:11,] %>%
  ggplot(aes(cc, str_wrap(str_trunc(datacentre, side = "right", 35), 15))) +
  geom_point(aes(size = n),
             colour =  wes_palette("Rushmore")[1],
             shape = 16) +
  theme_bw() +
  xlab("") +
  ylab("") +
  scale_size_continuous(range=c(7,25)) +
  geom_text(aes(label = prettyNum(n, big.mark = ","))) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

>>>>

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
keywords          : ""
wordcount         : "" # '`r  prettyNum(unname(stringi::stri_stats_latex(readLines("paper.Rmd"))[4]), big.mark = ",")` (includes reference list )'

@@

output:
  datacitation::apa6_word

abstract: "" # |
   # "How do archaeologists share their research data, if at all? We review what data are, according to current influential definitions, and previous work on the benefits, costs and norms of data sharing in the sciences broadly. To understand data sharing in archaeology, we present the results of three pilot studies: requests for data by email; review of data availability in published articles, and analysis of archaeological datasets deposited in repositories. We find that archaeologists are often willing to share, but discipline-wide sharing is patchy and ad hoc. Legislation and mandates are effective at increasing data-sharing, but editorial policies at journals lack adequate enforcement. Although most of data available at repositories are licensed to enable flexible reuse, only a small proportion of the data are stored in structured formats for easy reuse. We present some suggestions for improving the state of date sharing in archaeology, among these is a standard for citing data sets to ensure that researchers making their data publicly available receive appropriate credit."
---

```{r, setup2, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  eval = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)

@@

Data are the building blocks of research, and access to data is necessary to fully understand and extend the work of others (Fienberg et al 1985). Yet archaeological data are often not available for a variety of reasons, including a perceived lack of credit for making data available. The goal of this paper is to present a standard for the scholarly citation of archaeological data, akin to the way we cite publications. The need for this has been motivated by Huggertt’s (2017) findings of low levels of data citation and reuse in archaeology, and Clarke’s (2015) analysis that digital data preservation and management systems are under-developed in archaeology, despite our obligations as stewards of cultural heritage. A standard for data citation is important because it gives credit to the authors or compilers of the data, and so accords them visibility and recognition for their scholarly contribution in a way that is valued broadly in the discipline. A citation standard is also important for tracking the use of the data, and generating metrics about the impact of the dataset on the discipline. Understanding the use of resources such as datasets is important for allocating resources within archaeology and identifying novel research directions.

Archaeologists work with data in research, commercial, government, and other contexts. A full investigation of data in each of these contexts is beyond the scope of this paper, so we limit this study to what we are most familiar with, the research context. However, many of the practical details we discuss here will also be relevant to archaeologists working in other contexts. We begin by defining what we mean by ‘data’, and reviewing previous work on researchers’ attitudes to sharing data.  We motivate data sharing and citing behaviors by explaining why data sharing is beneficial to archaeology as a discipline, and to individual researchers. Then we survey some of the data sharing and citing norms that have emerged in other research areas. From these norms we derive a standard suitable for citation of archaeological data. Finally, we present some basic instructions on how to share archaeological data.

@@

Defining what data are is an ongoing research problem for information scientists (Leonelli 2016; Borgman 2012). Table \@ref(tab:summary-of-defs-of-data) summarizes some of the definitions debated in the philosophy of science literature. A more concrete starting point is this definition from the National Academies of Sciences:

@@

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different challenges and requirements for access, archiving, and so on. The categories shown in Table \@ref(tab:types-of-data) are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: "What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models." These definitions in policy are consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependent on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). How can these definitions and philosophical discussions inform our practice? They tell us that what counts as data is emergent from our professional interactions. Data are the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email. They are not exclusively objective archaeometric records, but include interpretive and derivative results. In short, data are anything another person needs to understand our work better.

To translate this into a more applied context, our practice of data sharing it to make available the computer files (e.g. containing text, raw structured data, images, etc.) that we analyze or otherwise use to create the figures, tables and quantitative results that we include in our reports and publications. These include our derived results and secondary data, such as taxonomic and typological classifications. This selection omits materials that could also be considered data according to the definitions noted above (e.g. field notes), but we feel our selection represents the most information-dense and meaningful items that are useful for our community of interest. These are also the items that have been most frequently requested from us by others in our professional interactions. Our citation standard is inclusive and imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion.

@@

While public and community benefits and costs are well-rehearsed and widely agreed on (Tables \@ref(tab:sharingcosts) and \@ref(tab:sharingbenefits)), they offer no, or only weak, tangible incentives to the individual researcher (Rowhani-Farid et al. 2017). Previous work has identified citation advantages as a tangible benefit to individual researchers who share data with publications, although the magnitude of the advantage varies greatly (see Table \@ref(tab:indivbenefits) for a summary). Increased publication productivity also appeared to be associated with data sharing behaviors. Across over 7,000 NSF and NIH awards, Pienta et al. (2010) reported that research projects with archived data produced a median of 10 publications, versus only five for projects without archived data.

```{r sharingcosts}
tbl <- readr::read_csv("../figures/tbl_reasons_not_to_share_data.csv")
tbl[,2] <- ""
names(tbl) <- c(names(tbl)[1], "")
knitr::kable(tbl, caption = "Summary of public and community costs of data sharing, from Stodden (2010) and Borgman (2012)")
```

.

```{r sharingbenefits}
tbl <- readr::read_csv("../figures/tbl_reasons_to_share_data.csv")
tbl[,2] <- ""
names(tbl) <- c(names(tbl)[1], "")
knitr::kable(tbl, caption = "Summary of public and community benefits of data sharing, from Stodden (2010) and Borgman (2012)")
```

.

@@

Stodden (2010) surveyed registrants at a top machine learning conference (the Neural Information Processing Systems (NIPS) conference) to identify the factors that affect researchers’ decisions to reveal data (Table \@ref(tab:costsbenefits)). Stodden found that communitarian factors influence scientists to share their data, but private incentives determine when scientists choose not to share their code and data. In the top three reasons not to share, we see a lack of incentives to invest time in data sharing (however, there are many other time-consuming tasks in research, and they get done because clear reward structures are in place); a lack of faith in data citation (which we hope this paper will help with); and a lack of an efficient copyright process. Stodden’s (2009a, b) ‘Reproducible Research Standard’, summarized in Table \@ref(tab:stodden-rrs), is a convenient and robust copyright schema that is in wide use for many typical research projects.

@@

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how do these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviors and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). This intensive-restricted spectrum also maps on to Sawyer's (2008) analysis of data sharing in data-rich fields compared to data-poor fields. Data rich fields, such as astronomy, oceanography, and ecology, are have three common characteristics: pooling and sharing data resources is expected; the cumulative nature of work that builds on common data and common methods reduces the likelihood of alternative data sources and other methods being common; and the number of theoretical choices is limited because there is enough data to adequately test competing theories. Data-poor fields have three common characteristics: data are rare, or hard to get, or both; the types of data available often dictate the methods taken; and theories are relatively easy to generate and more difficult to validate. In data-poor fields, researchers often invest substantial time and resources on gathering their own data. The getting of data, and the quality of the data collected, is often a key determinant of the status of a researcher, leading to hoarding of data as an asset to be traded for professional benefit.

Archaeologists can be found all over this spectrum because of the diversity of archaeological research. However, our observations of data sharing in archaeology, and the nature of archaeological research generally, is that it is, for the most part, currently a restricted data-sharing and a data-poor field. There are few standardized and widely used methods for sharing and citing data, a challenge for archaeological data sharing that we hope this paper will help to address.

Even though there is a wide spectrum of data sharing behaviors, overall rates of data sharing by individual researchers are very low in many fields. Wicherts et al. (2006) emailed authors of 141 empirical articles recently published by the American Psychological Association (APA) journals to request a copy of the original data reported in the articles. They received 64 of the 249 datasets they requested, resulting in a data sharing rate of about 26%, despite all the authors of the articles in their study being bound to the APA's Ethical Principles, which state authors must ‘not withhold the data on which their conclusions are based from other competent professionals who seek to verify the substantive claims’ (American Psychological Association, 2001, p. 396). Valpaemel et al. (2015) similarly requested data from 394 papers, published in all issues of four APA journals in 2012 and found that 38% of authors responded with their data. They concluded that although the data sharing rate increased as compared to the study by Wicherts et al., their findings are ‘worrisome’ and indicate a 'poor availability of data'.

@@

The absence of a standardized and sustainable approach to data sharing is starkly demonstrated by Vines et al.’s (2014) study of data availability over time in biology publications. They examined the availability of data from 516 studies of plants or animal morphology between 2 and 22 years old. Vines et al. found that odds of a data set referred to in a publication being available fell by 17% per year. They found that the key factors responsible for the unavailability of data was that the data were lost or on inaccessible storage media.

@@

Here we present the results of three small pilot studies aimed at understanding how archaeologists share their data. First, we emailed a sample of authors of research publications to request access to the data behind their publication. Second, we surveyed a sample of journal articles to see what data was available. Third, we analyzed information about archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The small sample sizes of the first two of these pilot studies limit how generalisable the results are, but as the first of their kind for archaeology, they are valuable for proposing hypotheses for larger-scale investigations.

@@

Our sample focused on _Journal of Archaeological Science_ articles on experiments with stone artefacts that were published during `r min(email_survey$'publication year', na.rm = TRUE)` and `r max(email_survey$'publication year', na.rm = TRUE)`. We focused on stone artefact experiments because all research materials in these projects are generated by the researcher, so there are no other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. Because we contacted authors of papers at the intersection of (1) containing eminently sharable data and (2) our _bona fide_ research interests (i.e. stone artefacts for BM), this is a targeted, non-random sample that may not be representative of the entire archaeological community. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymized so the specific authors we contacted are not identifiable.

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and received `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We received `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We found no effect on the author's student status at time of publication on data sharing behaviors, with roughly equal proportions of students and non-students sharing and not sharing.

Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014). Our finding is not unique - Chambers' (2017) examined 50 PLOS One papers on brain images published in November 2014 - May 2015 and found that only 38% of these papers archived their data in an open repository. Similarly, Nuijten et al. (2017) found that of 462 PLOS One articles that promised data availability, 29% of these did not have any available data. Unfulfilled promises to share data are not unique to PLOS One (Kidwell et al. 2016), but these studies highlight the need to resolve ambiguity among authors about what counts as data, and for editors and peer reviewers to more critically assess whether an author has made their data openly available.

We see similar rates of sharing in response to email requests among archaeologists to what have been described in other disciplines (Table \@ref(tab:sharing-rates), Tenopir et al. 2015; Vanpaemel et al. 2015). However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses we received suggests two key reasons why authors are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data (a typical concern of researchers in data-poor fields), and that their data are not in a form suitable for sharing and the task of organizing those data is not a high priority. These reasons not to share have also been documented in other research areas, noted above. The absence of patterns in the date of the paper, student status of the author and other variables (Figure \@ref(fig:email-figs)) may relate to more to our small sample size than a real absence of relationships in data sharing behavior.

@@

```{r student, fig.cap="No relationship between student status and data sharing behavior"}
sharing_by_student <-
email_survey %>%
  filter(!is.na( `email sent date`)) %>%
    mutate(`content recieved date` = if_else(!is.na(`content recieved date`),
  "yes", "no")) %>%
  group_by(
  `author is graduate student at time of publication?`,
  `content recieved date`) %>%
  tally() %>%
  filter(!is.na(`author is graduate student at time of publication?`)) %>%
  ggplot(aes(`author is graduate student at time of publication?`, n)) +
    geom_col(aes(fill = `content recieved date`),
             position = position_dodge(preserve = 'single')) +
  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3)],
                    guide = FALSE) +
  scale_y_continuous("Number of papers",
                     breaks= pretty_breaks()) +
    xlab("Author was graduate student at time of publication?") +
    theme_bw() +
  coord_fixed(ratio = 0.05)
```

@@

```{r journals, fig.cap="No relationship between sharing behavior and journal."}

@@

In addition to our email survey, we also conducted a pilot study of a random sample of `r nrow(jas_article_survey)` 48 articles published during February-May 2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviors. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data) (Figure \@ref(fig:data-types-and-files)). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, an open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

The _Journal of Archaeological Science_ has had a 'data disclosure' policy since at least 2013 (Editors, 2013) which states that "All data relating to the article must be made available in Supplementary files or deposited in external repositories and linked to within the article". The data availability rate of `r data_are_avail_perc`% reported in our pilot study reflects either weak enforcement of this policy or an incomplete understanding among editors, reviewers, and authors of how to interpret and implement this policy. The prominence of compositional and dating data among the shared data types suggests a least effort strategy by authors sharing data that do not require extensive cleaning and tidying (because the data are generated by instruments rather than entered by hand, such as artefact attribute data). The size of compositional and dating datasets means that they mostly fit easily into tables in the text or supplementary files, and do not require much effort to organize (e.g. very little metadata is required for others to use these data). The small proportion of papers in our sample using or sharing a scripting language suggests that unlike other fields (Eglen et al. 2017), archaeologists do not appear to recognize code for data analysis as a primary research object to share with the research community. Code written by the researcher for even the simplest operations should be shared along with data because code communicates decisions made about the data analysis that can help others understand, analyze, and reuse the data.

@@

plot_datacentre_rights_table <-
datacentre_rights_table[1:11,] %>%
  ggplot(aes(cc, str_wrap(str_trunc(datacentre, side = "right", 35), 15))) +
  geom_point(aes(size = n),
             colour =  wes_palette("Rushmore")[1],
             shape = 16) +
  theme_bw() +
  xlab("License") +
  ylab("Data repository") +
  scale_size_continuous(range=c(7,25)) +
  geom_text(aes(label = prettyNum(n, big.mark = ","))) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
